---
title: "LakeEnsemblR.WQ: Lake Mendota test case"
author: "TBO"
date: "2026-08-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")
```

## Purpose

An end-to-end, reproducible workflow for LakeEnsemblR.WQ on the Lake Mendota
test case (single inflow):

1. Export the coupled model inputs.
2. Validate the model setups.
3. Run the model ensemble.
4. Compute harmonized metrics.
5. Sensitivity analysis, then calibration (LHC, optional DE, write-back, and
   calibrating several coupled models at once).
6. Compare and visualize model outputs against each other and observations.

**How to run it:** knit or run this file from inside the test-case folder (the
one that contains `LakeEnsemblR.yaml`, `LakeEnsemblR_WQ.yaml`, `Output.yaml`,
the model folders and `standart_observed_data.csv`). No `setwd()` is needed
because knitr uses the Rmd's own folder.

## Prerequisites

- A physical LakeEnsemblR setup (`LakeEnsemblR.yaml` and its input files).
- A water-quality config file, `LakeEnsemblR_WQ.yaml`, and a metrics config
  file, `Output.yaml`.
- Model folders for the models you run: `GLM-AED2`, `GOTM-WET`,
  `GOTM-Selmaprotbas`, `Simstrat-AED2`.
- For calibration: an observed-data CSV with columns `datetime`, `depth`,
  `variable_global_name`, `value`.

```{r packages}
library(LakeEnsemblR)    # remotes::install_github("tubabucak/LakeEnsemblR")
library(LakeEnsemblR.WQ) # remotes::install_github("tubabucak/LakeEnsemblR.WQ")
library(dplyr)
library(tidyr)
library(ggplot2)

# Which optional model packages are installed? A missing one only matters if
# you run the matching model.
sapply(c("GLM3r", "WETr", "SelmaprotbasR", "SimstratR", "DEoptim"),
       requireNamespace, quietly = TRUE)
```

### Run settings

`quick = TRUE` uses tiny sample sizes so the whole document finishes fast and
proves the workflow works. Set `quick = FALSE` for a real calibration. Rule of
thumb for DE: `de_popsize` should be roughly 10x the number of calibrated
parameters (see `?DEoptim::DEoptim.control`).

```{r settings}
setwd(dirname(rstudioapi::getSourceEditorContext()$path))
quick    <- TRUE
obs_file <- "standart_observed_data.csv"   # datetime, depth, variable_global_name, value
models   <- c("GLM-AED2", "SIMSTRAT-AED2", "GOTM-Selmaprotbas", "GOTM-WET")

n_cores <- max(1L, min(4L, parallel::detectCores(logical = FALSE) - 1L))
cfg <- if (quick) {
  list(lhc_seq = 4,   lhc_par = 8,   workers = min(2L, n_cores),
       de_pop = 4,    de_iter = 2,   sens_steps = 5)
} else {
  list(lhc_seq = 50,  lhc_par = 300, workers = n_cores,
       de_pop = 60,   de_iter = 30,  sens_steps = 10)
}

set.seed(123)   # Latin hypercube sampling and DE are random
```

## 1) Export WQ configuration and inputs

```{r export}
export_config("LakeEnsemblR.yaml", model = c("GLM", "GOTM", "Simstrat"))
export_config_wq("LakeEnsemblR_WQ.yaml")
```

## 2) Validate model setups

`run_ensemble_wq()` can do this itself with `validate = TRUE`. Calling the
validators directly surfaces configuration errors before you commit to a full
run.

```{r validate}
validate_glm_aed(sim_folder = "GLM-AED2", file = "glm3.nml", verbose = TRUE)
validate_gotm_wet(sim_folder = "GOTM-WET", file = "gotm.yaml", verbose = TRUE)
validate_gotm_wet(sim_folder = "GOTM-Selmaprotbas", file = "gotm.yaml", verbose = TRUE)
validate_simstrat(sim_folder = "Simstrat-AED2", file = "simstrat.par", verbose = TRUE)
```

## 3) Run ensemble simulations

```{r run-ensemble}
run_res <- run_ensemble_wq(
  config_file = "LakeEnsemblR_WQ.yaml",
  models      = models,
  folder      = ".",
  validate    = TRUE,
  verbose     = TRUE
)
```

## 4) Compute metrics

Run this again after calibration to compute the harmonized metrics with the
calibrated parameters (see Section 5.6).

```{r metrics}
metric_out <- cal_metrics(
  metric_yaml_file = "Output.yaml",
  model_filter     = "all",
  wq_config_file   = "LakeEnsemblR_WQ.yaml"
)
```

## 5) Calibration workflow

### 5.1 Generate calibration tables

```{r calib-tables}
create_calibration_tables(
  folder         = ".",
  config_file    = "LakeEnsemblR_WQ.yaml",
  folder_out     = "calibration",
  models_coupled = models,
  bounds_factor  = 0.2
)
```

Edit the module files in the `calibration` folder:

- set `include = TRUE` for the parameters you want to calibrate
- adjust `lower`, `upper` and `initial` as needed

### 5.2 Build `calib_setup` from the edited tables

```{r calib-setup}
cs_all   <- calib_setup_from_tables(folder_in = "calibration", model_coupled = models)
cs_selma <- subset(cs_all, model_coupled == "GOTM-Selmaprotbas")
param_selma <- unique(cs_selma$pars)
param_selma
```

### 5.3 Sensitivity analysis first

Find which parameters actually matter before spending compute calibrating
them. `output_mode = "raw"` needs only the model-native variable name (no
`wq_config_file`). `group_name` restricts the change to one phytoplankton
group.

```{r sensitivity}
run_sens <- function(param, vars, group_name = NULL, n_steps = cfg$sens_steps) {
  args <- list(param_name = param, calib_setup = cs_selma, yaml_file = "Output.yaml",
               model_dir = "GOTM-Selmaprotbas", n_steps = n_steps,
               model = "GOTM-Selmaprotbas", output_mode = "raw", vars = vars)
  if (!is.null(group_name)) args$group_name <- group_name
  do.call(run_sensitivity, args)
}

sens <- list(
  kc      = run_sens("kc",   "selmaprotbas_DO_mg"),                          # light extinction
  pvel    = run_sens("pvel", "selmaprotbas_DO_mg"),                          # phytoplankton settling
  r0_diat = run_sens("r0",   "total_chlorophyll_calculator_result", "diatoms")  # diatom growth only
)
```

`plot_sensitivity()` draws one grey line per parameter value, a blue band for
the 5-95% range across them, and the median. `sensitivity_to_long()` returns
the same data as a tidy data frame if you want your own plot or numbers.

```{r sens-plots}
plot_sensitivity(sens$kc,      depth = 5, ylab = "DO (mg/m3)", title = "kc -> DO at 5 m")
plot_sensitivity(sens$pvel,    depth = 2, ylab = "DO (mg/m3)", title = "pvel -> DO at 2 m")
plot_sensitivity(sens$r0_diat, depth = 2, ylab = "Chl-a",      title = "diatom r0 -> total chlorophyll at 2 m")
```

A parameter whose envelope is a thin line has little effect on that variable
and is a poor calibration candidate. Wide envelopes mark the parameters worth
calibrating.

### 5.4 Latin hypercube (LHC) calibration, one model

`calib_wq()` samples the parameter space with Latin Hypercube Sampling, edits
the model's own config files in `model_dir`, runs the model and scores every
sample against the observations. It returns one row per sample and observed
variable with NSE / RMSE / NRMSE / PBIAS / KGE. The model's config files are
restored to their pre-calibration state when it returns, even on error.

```{r lhc}
res_lhc <- calib_wq(
  model          = "GOTM-Selmaprotbas",
  calib_setup    = cs_selma,
  param_names    = param_selma,
  yaml_file      = "Output.yaml",
  model_dir      = "GOTM-Selmaprotbas",
  n_samples      = cfg$lhc_seq,
  wq_config_file = "LakeEnsemblR_WQ.yaml",
  obs_file       = obs_file,
  target_variables = "DO_gramsPerCubicMeter",
  best_metric    = "KGE",
  verbose        = TRUE
)

head(res_lhc)
attr(res_lhc, "best_parameter_set")   # winning parameter set
```

### 5.5 Parallel LHC + Differential Evolution (DE) refinement

`parallel = TRUE` spreads the LHC samples over worker processes. `use_de =
TRUE` then runs DEoptim seeded with the best LHC samples. With DE,
`attr(res, "best_parameter_set")` is DE's refined optimum and the LHC-phase
best stays available under `attr(res, "lhc_best_parameter_set")`.

Keep `n_samples >= de_popsize`, otherwise the seeded DE population repeats
rows.

```{r lhc-de}
res <- calib_wq(
  model            = models,
  param_names      = param_selma,
  calib_setup      = cs_selma,
  yaml_file        = "Output.yaml",
  model_dir        = "GOTM-Selmaprotbas",
  obs_file         = obs_file,
  wq_config_file   = "LakeEnsemblR_WQ.yaml",
  target_variables = "DO_gramsPerCubicMeter",
  best_metric      = "KGE",
  stats_by_depth   = TRUE,
  return_best      = TRUE,
  verbose          = TRUE,
  save_results     = FALSE,
  parallel         = TRUE,
  n_workers        = cfg$workers,
  n_samples        = cfg$lhc_par,
  use_de           = TRUE,
  de_parallel      = TRUE,
  de_n_workers     = cfg$workers,
  de_popsize       = cfg$de_pop,
  de_iterations    = cfg$de_iter
)

attr(res, "best_parameter_set")       # DE-refined best (used for write-back below)
attr(res, "lhc_best_parameter_set")   # LHC-phase best, for comparison
```

**Where the logs are.** Worker processes have no console, so their output
goes to files inside the model folder: `GOTM-Selmaprotbas/lhc_debug_logs/lhc_workers.log`
(parallel LHC) and `GOTM-Selmaprotbas/de_debug_logs/de_workers.log` (DE).
Open them when a run looks stuck or every score is the penalty value.

### 5.6 Write the best parameters back

`write_best_calib_to_par_files()` writes the best row (DE's if `use_de =
TRUE`) into the module parameter files referenced by `LakeEnsemblR_WQ.yaml`,
or directly into the model config files with `write_target = "config"`. Note
that this **modifies your model files**: keep a copy if you want to return to
the defaults. Re-run `export_config_wq()` afterwards to push the values all
the way into the model folders.

```{r write-back}
write_best_calib_to_par_files(
  lhc_results  = res,
  calib_setup  = cs_selma,
  config_file  = "LakeEnsemblR_WQ.yaml",
  folder       = ".",
  metric       = "KGE",
  write_target = "config"
)
```

Re-run the calibrated model so its output reflects the new parameters:

```{r rerun}
run_ensemble_wq(
  config_file = "LakeEnsemblR_WQ.yaml",
  models      = "GOTM-Selmaprotbas",
  folder      = ".",
  validate    = TRUE,
  verbose     = TRUE
)
```

### 5.7 Check the calibration against observations

`plot_model_vs_obs_wq()` plots the simulation against the observations and
returns the plot and the fit statistics. For `GOTM-Selmaprotbas` and
`GOTM-WET`, zooplankton variables need `wq_config_file` so that the per-group
outputs can be summed.

```{r plot-vs-obs, fig.height=5}
# model, variable_global_name (must match obs_file's column), axis title
checks <- tibble::tribble(
  ~model,              ~variable,                              ~title,
  "GOTM-Selmaprotbas", "DO_gramsPerCubicMeter",                "DO (g/m3)",
  "GOTM-Selmaprotbas", "NO3_gramsPerCubicMeter",               "NO3 (g/m3)",
  "GOTM-Selmaprotbas", "NH4_gramsPerCubicMeter",               "NH4 (g/m3)",
  "GOTM-Selmaprotbas", "TP_gramsPerCubicMeter",                "TP (g/m3)",
  "GOTM-Selmaprotbas", "Total_Chla_miligramsPerCubicMeter",    "Chl-a (mg/m3)"
)

fit <- lapply(seq_len(nrow(checks)), function(i) {
  out <- plot_model_vs_obs_wq(
    config_file          = "Output.yaml",
    model                = checks$model[i],
    obs_data             = obs_file,
    variable_global_name = checks$variable[i],
    y_title              = checks$title[i],
    wq_config_file       = "LakeEnsemblR_WQ.yaml"
  )
  print(out$plot)
  out
})
names(fit) <- paste(checks$model, checks$variable, sep = " | ")
```

**Reading the statistics.** KGE, NSE and NRMSE are scale-free. RMSE is in the
model's own units, e.g. mg O2/m3 for Selmaprotbas (the observations are
converted with the dictionary `conversion_factor` when `obs_to_model_units =
TRUE`). PBIAS is `100 * sum(obs - sim) / sum(obs)`, positive = the model
underestimates, and it is not meaningful for variables that are often zero
(anoxic DO).

### 5.8 Calibrate all coupled models with `cali_ensemble_wq()`

`cali_ensemble_wq()` is the multi-model wrapper around `calib_wq()`. Pass the
full `models` vector and the combined `cs_all` table (split automatically by
`model_coupled`) and it runs, scores and optionally writes back the best
parameters for every model in one call.

Models run sequentially by default. `parallel_models = TRUE` runs one worker
process per model instead. It stacks with `parallel` / `de_parallel`, so size
the worker counts so that `n_model_workers * n_workers` does not exceed your
CPU cores. With `parallel_models = TRUE` progress is written to
`<output_dir>/cali_ensemble_wq_workers.log`.

```{r cali-ensemble}
ens_models <- "GOTM-Selmaprotbas"   # add more, e.g. c("GOTM-WET", "GOTM-Selmaprotbas")

result_all <- cali_ensemble_wq(
  models           = models,
  calib_setup      = cs_all,
  yaml_file        = "Output.yaml",
  folder           = ".",
  n_samples        = cfg$lhc_seq,
  n_workers        = cfg$workers,
  use_de           = TRUE,
  de_popsize       = cfg$de_pop,
  de_iterations    = cfg$de_iter,
  de_parallel      = TRUE,
  de_n_workers     = cfg$workers,
  parallel_models  = FALSE,
  obs_file         = obs_file,
  wq_config_file   = "LakeEnsemblR_WQ.yaml",
  ler_config_file  = "LakeEnsemblR.yaml",
  target_variables = "DO_gramsPerCubicMeter",
  best_metric      = "KGE",
  parallel         = TRUE,
  force_parallel_glm_simstrat = TRUE,
  verbose          = TRUE,
  on_error         = "skip",
  save_results     = TRUE,
  output_dir       = "calibration_results",
  write_best       = TRUE,
  write_target     = "config"
)

result_all$summary                                  # any failures or skips?
sapply(result_all$best_parameter_sets,              # final objective per model
       function(x) if (!is.null(x)) x$objective_value)
```

**DE convergence.** `bestvalit` is the best (minimized) objective per
generation. It can only stay flat or improve. If it is still falling at the
last generation, raise `de_iterations`. If it is flat from the start, the seeded
LHC starting point was already near the optimum, or `de_popsize` is too small
to explore.

```{r de-convergence, fig.height=4}
plot_de_convergence <- function(result) {
  for (m in names(result$results)) {
    de_obj <- attr(result$results[[m]], "de_phase")
    if (is.null(de_obj)) {
      message(m, ": no DE phase found (model failed, or DE was off)")
      next
    }
    bv <- de_obj$member$bestvalit
    plot(bv, type = "l", lwd = 2, xlab = "DE generation",
         ylab = "objective (minimized)", main = m)
    if (length(bv) >= 3) {
      total <- bv[1] - bv[length(bv)]
      tail2 <- bv[length(bv) - 2] - bv[length(bv)]
      message(sprintf("%s: total improvement %.4f, last 2 generations %.4f (%.0f%% of total)",
                      m, total, tail2, 100 * tail2 / total))
    }
  }
}
plot_de_convergence(result_all)
```

## 6) Compare and visualize model outputs

These plug directly into `cal_metrics()`'s nested list (Section 4) and the
NetCDF file from `create_netcdf_output()`.

### 6.1 Combine all models into one NetCDF

```{r netcdf}
metric_out <- cal_metrics(
  metric_yaml_file = "Output.yaml",
  model_filter     = "all",
  wq_config_file   = "LakeEnsemblR_WQ.yaml"
)

create_netcdf_output(
  output_lists    = metric_out,
  folder          = ".",
  model           = c("GLM", "SELMAPROTBAS", "WET", "SIMSTRAT"),
  ler_config_file = "LakeEnsemblR.yaml",
  wq_config_file  = "LakeEnsemblR_WQ.yaml",
  out_file        = "ensemble_output.nc"
)
```

### 6.2 Compare one metric across models

`compare_models_metric()` takes the `cal_metrics()` list directly.
`compare_models_metric_netcdf()` reads from the NetCDF instead, which is useful
for a variable that is not in `Output.yaml`.

```{r compare}
compare_models_metric(metric_out, metric = "Temp_degreeCelcius", depth = 1)$plot

```

### 6.3 Stratification, heatmaps, anoxia and ice

`plot_strat_metrics()`, `plot_anoxic_metrics()` and `plot_ice_metrics()` take
either the `cal_metrics()` list or a NetCDF path. `plot_heatmap_wq()` draws a
depth-time heatmap faceted by model; use any metric name present in
`Output.yaml`.

```{r strat-heatmap-anoxia}
strat <- plot_strat_metrics(metric_out, metric_name = "Duration_of_Stratification")
strat$duration_plot
strat$onset_plot
strat$mixing_plot

plot_heatmap_wq("output/ensemble_output.nc", "Temp_degreeCelcius", spin_up = 30)
plot_heatmap_wq("output/ensemble_output.nc", "DO_gramsPerCubicMeter", spin_up = 30)

plot_anoxic_metrics("output/ensemble_output.nc")
plot_ice_metrics("outputensemble_output.nc", metric_name = "Ice_Thickness_meter")
```

### 6.4 Lower-level time-series and scatter plots

`compare_plot()` and `scat_plot()` take one wide-format data frame per model
(`datetime` plus `Depth_<n>` columns, the shape `get_output_wq()` returns) and
one for the observations, and compute `cal_stats()` internally. See
`?compare_plot` and `?scat_plot` for the exact requirements.

## Session info

```{r session-info}
sessionInfo()
```
